Modelling BCG vaccination in the UK

### 2019-12-17 rmds/00-abstract.Rmd
remove UK born for children
finalised policy implication additions

Prior to the change in policy, several studies were carried out to assess the impact. I recreated one such study and found that there was a greater impact than previously thought. \par

@@

Statistical modelling is restricted by the available data. Therefore, I developed a dynamic model of TB, fit to available data, to forecast the impact of the policy change. Although the fit to the data was poor, the forecasts suggested that continuing school-age vaccination reduced TB incidence in the UK born compared with neonatal vaccination. Neonatal vaccination reduced incidence in UK born children but had little impact on other age groups.\par

>>>>

Prior to the change in policy, several studies were carried out to assess the impact. I recreated one such study, correcting a methodological flaw, and found that there was a greater impact than previously thought. \par

@@

Statistical modelling is restricted by the available data. Therefore, I developed a dynamic model of TB, fit to available data, to forecast the impact of the policy change. Although the fit to the data was poor, the forecasts suggested that continuing school-age vaccination reduced TB incidence in the UK born compared with neonatal vaccination. Neonatal vaccination reduced incidence in children but had little impact on other age groups.\par

### 2019-12-17 rmds/09-model-fitting.Rmd
finalised policy implication additions

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that underreporting is more likely than over-reporting. The prior for the reporting noise is based on the observed variation between years. This observation model is also used when incorporating non-UK born incidence rates into the models force of infection (Chapter \@ref(model-development)). This allows the uncertain in these observations to be properly accounted for in the incidence estimates produced by the fitted model.

@@

Several key data inputs such as incidence in the non-UK born population, coverage of the BCG vaccination program, births, deaths and the contact rate are not perfectly observed, or recorded, and may vary across time. For this reason these parameters are included into the model developed in the last chapter as noise terms. This means that they are resampled for each timestep and so vary stochastically over time. This results in a model that is semi-stochastic rather than being fully deterministic (Chapter \@ref(introduction)). A semi-stochastic model can be defined as a deterministic model that incoperates stochastic elements but that is still solved as a determinsitc system within a given timestep. It is a modelling approach that has been used prevously in the literature when key parameters are uncertain and potentially time varying.[@Funk2016a] For further details of the stochastic parameters included in the model see Chapter \@ref(model-development).

@@

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
           str_to_sentence() %>% 
           str_replace("-uk", "-UK")
    ) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit. Using 1000 samples from the posterior distribution of the fitted model for each scenario",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

>>>>

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that underreporting is more likely than over-reporting. The prior for the reporting noise is based on the observed variation between years. This observation model is also used when incorporating non-UK born incidence rates into the models force of infection (Chapter \@ref(model-development)). This allows the uncertainty in these observations to be properly accounted for in the incidence estimates produced by the fitted model.

@@

Several key data inputs such as incidence in the non-UK born population, coverage of the BCG vaccination program, births, deaths and the contact rate are not perfectly observed, or recorded, and may vary across time. For these reasons, these parameters are included in the model developed in the last chapter as noise terms. This means that they are resampled for each timestep and so vary stochastically over time. This results in a model that is semi-stochastic rather than being fully deterministic (Chapter \@ref(introduction)). A semi-stochastic model can be defined as a deterministic model that incoperates stochastic elements but that is still solved as a determinsitc system within a given timestep. It is a modelling approach that has been used prevously in the literature when key parameters are uncertain and potentially time varying.[@Funk2016a] For further details of the stochastic parameters included in the model see Chapter \@ref(model-development).

@@

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
           str_to_sentence() %>% 
           str_replace("-uk", "-UK")
    ) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit. Using 1000 samples from the posterior distribution of the fitted model for each scenario.",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

### 2019-12-17 rmds/98-discussion.Rmd
remove UK born for children
finalised policy implication additions

In Chapter \@ref(data), I explored TB epidemiology in England with a focus on BCG vaccination and data completeness. I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out), I used logistic regression to estimate associations between BCG vaccination and TB outcomes. I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland), I explored some of the modelling evidence that was used by policy makers to assess the impact of ending school-age universal BCG vaccination. I found that the previous approach had underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending universal school-age BCG vaccination was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff), where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations, using Possion and negative binomial models. However, in this chapter, I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Using this approach, I was unable to rule out an unrelated policy change as the cause of this reduction. Finally, in Chapter \@ref(model-results), I forecast the impact of various vaccination scenarios using a dynamic TB model that was developed and fitted in Chapter \@ref(model-development) and Chapter \@ref(model-fitting). I found that the BCG schools scheme was projected to reduce UK born TB incidence compared to both neonatal and no vaccination over a range of time horizons. However, neonatal vaccination reduced incidence in UK born children compared to any other scenario, although it had little impact in any other age group. No vaccination programme evaluated had an impact on incidence in older adults. The results from Chapter \@ref(model-results) are preliminary as the model on which they were based was a very poor fit to the observed data (Chapter \@ref(model-fitting)).

@@

>>>>

In Chapter \@ref(data), I explored TB epidemiology in England with a focus on BCG vaccination and data completeness. I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out), I used logistic regression to estimate associations between BCG vaccination and TB outcomes. I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland), I explored some of the modelling evidence that was used by policy makers to assess the impact of ending school-age universal BCG vaccination. I found that the previous approach was methodologically flawed and had underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending universal school-age BCG vaccination was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff), where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations, using Possion and negative binomial models. However, in this chapter, I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Using this approach, I was unable to rule out an unrelated policy change as the cause of this reduction. Finally, in Chapter \@ref(model-results), I forecast the impact of various vaccination scenarios using a dynamic TB model that was developed and fitted in Chapter \@ref(model-development) and Chapter \@ref(model-fitting). I found that the BCG schools scheme was projected to reduce UK born TB incidence compared to both neonatal and no vaccination over a range of time horizons. However, neonatal vaccination reduced incidence in children compared to any other scenario, although it had little impact in any other age group. No vaccination programme evaluated had an impact on incidence in older adults. The results from Chapter \@ref(model-results) are preliminary as the model on which they were based was a very poor fit to the observed data (Chapter \@ref(model-fitting)).

@@

Policy makers in the UK in particular should consider the implications for BCG vaccination of the results from this thesis. Of particular importance is the finding that the previously published approach for estimating the impact of ending BCG vaccination in the UK was methodologically flawed, spuriously precise and, when updated with newly available data, produced much larger estimates of the impact of ending BCG vaccination than previously thought. This finding highlights the importance of independent validation of modelling studies by subject area experts. To counterbalance this, the finding that overall the change in BCG vaccination policy was associated with decreased incidence rates indicate that the change in policy may have been justified. However, the benefit of the policy change was focussed on the non-UK born population who may have been impacted by other policy changes. Considering UK born cases alone, this thesis provides evidence that TB cases increased, with cases decreasing in young children. The modelling evidence, although preliminary, supported the finding that the change in policy would lead to increased overall incidence rates in the UK born population but reduce incidence rates in young children compared to school-age vaccination. On top of these specific findings, the points made in the previous paragraph also apply to the UK.

Unfortunately, definite recommendations cannot be made to policy makers based on the findings from this thesis. This is due to several limitations of the evidence and due to the inherent complexity of both TB and the BCG vaccine. A particular limitation is the difficulty in identifying who - post the change in BCG vaccination policy - should have been vaccinated and who should not have been. This meant that targeted high-risk neonatal vaccination could not be modelled and therefore could not be compared to school-age BCG vaccination. It also meant that population-level studies had to be used when studying the impact of change in policy. As the impact of the targeted scheme was likely focussed on the groups it targeted this may have diluted the impact of this policy. An additional issue is that whilst a trade-off was identified between reducing the number of TB cases in young children at the cost of a larger number of adult TB cases the impact of TB symptoms on children compared to adults was not considered. To be able to make more precise recommendations additional data sources are required. The first of these is a measure of the impact of TB on individuals at various ages, such as a quality-adjusted life-year (QALY). This would ideally be arrived at by assessing expert opinion and could then be used to identify what level of trade-off between childhood and adult TB cases resulting in the lowest overall QALY burden. Finally, data is required on who should have been vaccinated under the targeted high-risk neonatal vaccination scheme and of these individuals how many of them then went on to develop TB. This data would allow more precise statistical estimates to be made of the overall impact of the change in policy and would also enable targeted high-risk neonatal vaccination to be included in the dynamic TB model. Inclusion in the dynamic model would then allow the impact of various targeted vaccination scenarios to be considered over an extended time-frame and compared to universal vaccination scenarios.
